% readme.tex -- a short example of how each Stata Journal insert should be
% organized.

\inserttype[st0001]{article}
\author{Michler, Josephson, Barriga Cabanillas, Michuda, Oliver}{%
  Jeffrey D. Michler\\University of Arizona\\Tucson, AZ\\jdmichler@arizona.edu
  \and
  Anna Josephson\\University of Arizona\\Tucson, AZ\\jdmichler@arizona.edu
  \and
  Oscar Barriga Cabanillas\\World Bank\\Washington, D.C.\\obarriga@worldbank.org
  \and
  Aleksandr Michuda\\Swarthmore College\\Swarthmore, PA\\amichud1@swarthmore.edu
  \and
  Jeffrey C. Oliver\\University of Arizona\\Tucson, AZ\\jdmichler@arizona.edu
}
\title[wxsum]{wxsum: A command for processing temperature and precipitation data}
\maketitle

\begin{abstract}
In this article, we introduce the community-contributed command \texttt{wxsum} which processes daily temperature and precipitation data and outputs useful summary statistics as well as measures of weather ``shocks.'' The command calculates 18 different measures of precipitation and 13 different measures of temperature. These include simple summary statistics, like the mean and standard deviation, as well as measures of shocks, such a z-score, dry spells, and killing degree days (KDD). The goal of the command is to reduce the cost of using weather data in socioeconomic analysis by allowing for quick and accurate calculations of the most common weather metrics used in the literature.

\keywords{\inserttag, wxsum, Earth observation, weather data, summary statistics}
\end{abstract}

% discussion of the Stata Journal document class.
\input wxsum_body_v1.tex
% discussion of the Stata Press LaTeX package for Stata output.
%\input stata.tex

\bibliographystyle{sj}
\bibliography{wxsum_ref_v1}

\begin{aboutauthors}
Jeffrey D. Michler is an Associate Professor in the Department of Agricultural and Applied Economics at the University of Arizona.

Anna Josephson is an Associate Professor in the Department of Agricultural and Applied Economics at the University of Arizona.

Aleksandr Michua is a Visiting Assistant Professor at the Department of Economics at Swarthmore College.

Oscar Barriga Cabanillas is an Economist at the World Bank.

Jeffrey C. Oliver is a Data Science Specialist in the University of Arizona Libraries.
\end{aboutauthors}

\endinput
